# Exploratory Data Analysis

## Overview{-}

This session has 4 components:

1. Data, data summaries and generic plotting in R (this should be refresher!)
2. An introduction to visualisation using the `ggplot2` package (this builds on its surreptitious introduction in Week 1!)
3. Data wrangling with `dplyr`
4. Using piping syntax

There is also an optional additional section on enhanced mapping with the `ggplot2` package (loaded with the `tidyverse` package). 

The overarching theme of this practical is *Explanatory Data  Analysis* or EDA. The aim of EDA is to generate understandings of the data, by revealing patterns, trends and relationships within and amongst the data variables. EDA can involve numeric summaries of data properties (data distribution, central tendencies, spread, etc), correlations with other variables, summarising these in tables, or EDA can summarise data graphically, through the use of different types of plots. 

EDA in this way provides information about the *structure* of the data - the patterns, trends, interactions and relationships within and between variables. It is a critical part of any data analysis. It underpins the choice and development of methods and approaches of analysis, including which variables to include in modelling.

You will need to load some  packages. Note you will probably need to install the `RColorBrewer` package. **Remember** that packages should be installed with their dependencies, and they only need to installed once, and after they are installed they can simply be called using the `library` function. So you only need to install packages that you have not installed before in earlier weeks. Last week you should have installed `tidyverse` and `sf` in the main practical. 

The code below lists the packages in an object called `packages` and then tests to see of they installed with the result stored in `not_installed` and then installs missing packages:

```{r eval = F}
# check and install missing packages
packages <- c("sf", "RColorBrewer", "tidyverse")
# check which packages are not installed
not_installed <- packages[!packages %in% installed.packages()[, "Package"]]
# install missing packages
if (length(not_installed) > 1) {
  install.packages(not_installed, repos = "https://cran.rstudio.com/", dep = T)
}
```


```{r eval = T, warning = F, message = F}
library(tidyverse)
library(RColorBrewer)
library(sf)
```



## Data, data summaries and generic plotting in R

The data used in this session are area-level demographic and voting data covering the UK’s referendum vote on membership of the EU. You’ll start by exploring the results data published for all 380 Local Authority Districts (LADs) in Great Britain (GB) made available in by The Electoral Commission. Later you’ll explore the geography of this vote against demographic data collected from the 2011 Census. These data have been collated for these practicals into a single spatial dataset in `geojson` format (`data_gb.geojson`). 

```{r echo = T, eval = T, message=FALSE, warning=F}
data_gb <- st_read("data_gb.geojson", quiet = T)  
```

The `data_gb` object you have loaded is in `sf` format in R. You can examine the data and the attributes that are included:
```{r eval = F}
data_gb
names(data_gb)
summary(data_gb)
plot(st_geometry(data_gb))
st_bbox(data_gb)
```

The aim of EDA is to understand data properties and data structure. 

R contains many functions for examining **numeric** data properties and the dollar sign (`$`) can be used to access individual variables (attributes):

```{r eval = F}
summary(data_gb$share_leave)    # summary
mean(data_gb$share_leave)       # mean
sd(data_gb$share_leave)         # standard deviation
var(data_gb$share_leave)        # variance
IQR(data_gb$share_leave)        # interquartile range
```

These can be called for individually named variables as above. Some generic plotting of continuous variables can also be used. Note that the density plot is a smoothed version of the histogram. There is a link at the end of this practical that has full explanation of density histograms. 

```{r eval = F}
# histogram
hist(data_gb$share_leave)
# denisty histogram
hist(data_gb$share_leave, prob=T)   
# add a line
lines(density(data_gb$share_leave,na.rm=T),
      col='olivedrab',lwd=2)
# and ticks indicating the data points (it looks like a rug!)
rug(data_gb$share_leave)
```

And boxplots can be called in the same way
```{r eval = F}
boxplot(data_gb$share_leave, main = "Leave Share by LADs")    
```

Recall that histograms are visual representation of the data / variable distributions, and boxplots show the median and inter-quartile ranges - some further information on these is available here:

- https://www.khanacademy.org/math/statistics-probability/summarizing-quantitative-data/box-whisker-plots/a/box-plot-review
- https://wellbeingatschool.org.nz/information-sheet/understanding-and-interpreting-box-plots

And different groups can be compared, here grouping by Region:
```{r ch2fig1, fig.height = 7, fig.cap = "Boxplots of the leave vote share, grouped by regions." }
# save the plot parameters and set new ones
old.par = par()
par = par(mar = c(12,4,4,4))
# do the boxplots by groups and shade
boxplot(data_gb$share_leave ~ data_gb$Region, 
        las = 2, xlab = "", ylab = "Vote leave share", 
        col = brewer.pal(11, "Spectral"))
# reset the plot parameters
par = old.par
```

Non-numeric data can also be summarised using tables and barplots. 
```{r eval = F}
# table
table(data_gb$Region)
# xtabs for 1 variable
xtabs(data_gb$Remain ~ data_gb$Region)
# xtabs for 2 variables summarised by region 
xtabs(cbind(Remain = data_gb$Remain, Leave = data_gb$Leave) ~data_gb$Region)
```
Barplots require summarised data, hence the creation of the `tab` variable below:
```{r eval = F}
# create a table
tab = xtabs(data_gb$Leave ~ data_gb$Region) / 1000
# call the bar plot (note the use of sort!)
par = par(mar = c(12,4,4,4))
barplot(sort(tab), las = 2, 
        col =  brewer.pal(11, "Spectral"),
        main = "Leave votes (1000s)")
par = old.par
```

The above code snippets have shown the default plotting for single numeric variables, and counts of categorical ones, with some grouping (in this case by Regions). The final set of plotting routines examines the interaction of 2 numeric variables using a scatterplot. 

The code below summarises the correlations between variables and then shows these graphically:

```{r ch2fig2, eval = T, fig.cap = "A simple scatterplot."}
# correlation
cor(x = data_gb$share_leave, y = data_gb$degree_educated)
# scatterplot
plot(x = data_gb$share_leave, y = data_gb$degree_educated)
```

This can be controlled and enhanced (examine the help for points - `?points`):
```{r ch2fig3, eval = T, fig.cap = "A simple scatterplot with a trend line."}
plot(degree_educated ~ share_leave, data = data_gb,
     pch = 20, col = "tomato",
     xlab = "Leave Share", ylab = "Prop. Degree educated")
# add a trend line
abline(lm(degree_educated ~ share_leave, data = data_gb))
```

## An introduction to visualisation using the `ggplot2` package

The `ggplot2` package supports many types of visual summary. It allows multiple, simultaneous visualizations, supports grouping by colour, shape and facets. The precise choice of which, out of  the many possible options, requires some careful thinking. Does the data need to be sorted or faceted? Should transparency be used. Do I need to group the data? etc. 

The `ggplot2` package is installed with the `tidyverse` package. R has several systems for making visual outputs such as graphs, but `ggplot2` is one of the most elegant and most versatile. It implements the **grammar of graphics** [@wilkinson2012grammar], a coherent system for describing and constructing graphs using a *layered* approach. If you'd like to learn more about the theoretical underpinnings of `ggplot2`, "The Layered Grammar of Graphics" [@wickham2010layered] is recommended.

### `ggplot` basics
The basic idea is that graphs are composed of different layers each of which can be controlled. The basic syntax is: 
```{r eval = F}
# do not run this code - it is an illustration! #
# specify ggplot with some mapping aesthetics
ggplot(data = <data>, mapping = <aes>) +
  geom_<function>()
```
In this syntax first `ggplot` is called, a dataset is specified (`<data>` in the above) and some mapping aesthetics are defined (`<aes>`). Together, these tell `ggplot` which variables are to be plotted (for example, on the $x$ and $y$ axes), which variables are to be grouped and coloured. Finally, `geom_<function>()` specifies the plot type such as `geom_point()`, `geom_line()` etc. It has a default set of plot and background colours, line types and sizes for points, line width, text etc, all of which can be overwritten. 

These basics are probably best described through illustration. The code below recreates the previous scatterplot and assigns it to `p`:

```{r eval = F}
p = ggplot(data = data_gb, aes(x = share_leave, y = degree_educated)) + 
    geom_point(col = "dodgerblue", shape = 20) +
    xlab("Leave Share") + ylab("Prop. Degree educated")
```

This can then be called to generate the plot
```{r eval = F}
p
```

However because of the layered nature it can also be augmented and further graphical elements can be added as layers. The example below fits a linear trend line:
```{r eval = F}
p + geom_smooth(method = "lm")
```

This is equivalent to:
```{r eval = F}
p = ggplot(data = data_gb, aes(x = share_leave, y = degree_educated)) + 
    geom_point(col = "dodgerblue", shape = 20) +
    xlab("Leave Share") + ylab("Prop. Degree educated") +
    geom_smooth(method = "lm")
p
```

:::: {.noticebox}
**Note** that there is some flexibility in the `ggplot` syntax. An alternative is to specify the data with the plot function:
```{r eval = F}
ggplot() +
  geom_<function>(data = <data>, mapping = <aes>)
```

This syntax is useful if you want to plot different layers or objects together. 
::::


Histograms and boxplots of single continuous variables can be visualised and for histograms the `bins` parameter controls the number of histogram groups (columns):

```{r eval = F}
# a density plot
ggplot(data = data_gb, aes(x = share_leave)) +
  geom_density()
# a histogram
ggplot(data = data_gb, aes(x = share_leave)) +  
  geom_histogram(bins = 30, col = "red", fill = "salmon")
# a boxplot
ggplot(data = data_gb, aes(x = share_leave)) +
  geom_boxplot(fill = "tomato", width = 0.2) + 
  xlab("Leave Share") + ylab("")
```

A *density histogram* can also be constructed by specifying a density aesthetic for `y` in the `geom_histogram` parameter `after_stat(density)`. The areas of each bin indicate the relative probability of the bins in the distribution (the sum of the bin counts is 1). A density function can also be added to the histogram using the code below:

```{r ch2fig4, eval = T, fig.cap = "A density histogram with a density plot."}
ggplot(data_gb, aes(x = share_leave)) +  
  geom_histogram(aes(y=after_stat(density)),bins = 30, 
                 fill="tomato", col="white") +
  geom_density(alpha=.5, fill="#FF6666")
```

The line `geom_histogram(aes(y=..density..)...` rescales the histogram counts so that bar areas integrate to 1, but is otherwise the same as the standard `ggplot2` histogram. The `geom_density()` function adds  a density curve to the plot with a transparency term (`alpha=.5`). You could plot just the density curve by running the code below.
```{r eval = F}
ggplot(data_gb, aes(x = share_leave)) +	
  geom_density(aes(y=after_stat(density)), alpha=.5, fill="#FF6666")
```

The `ggplot2` package includes a number of pre-defined styles, called by `theme_X()`, that can be added as a layer. The code below applies `theme_minimal()` as in the figure below. 
```{r eval = F}
ggplot(data_gb, aes(x = share_leave, y = degree_educated)) + 
  # specify point characteristics
  geom_point(col = "tomato", shape = 20, size = 2.7) +
  xlab("Leave Share") + ylab("Prop. Degree educated") +
  # specify a trend line and a theme/style
  geom_smooth(method = "lm", colour = "#DE2D26") +
  theme_minimal() 
```

You should explore other `ggplot` themes which are listed if you enter `?theme_bw` at the console.

In a similar way bar plots can be constructed. The code below does this but `ggplot2` does not need tabulated data to be passed to it:

```{r eval = F}
unique(factor(data_gb$Region))
ggplot(data_gb, aes(y = factor(Region))) + 
  geom_bar(stat = "count") + xlab("Count")
```
This can be ordered or sorted using the `fct_infreq` function from the `forcats` package loaded with tidyverse:
```{r eval = F}
ggplot(data_gb, aes(y = fct_infreq(factor(Region)))) + 
  geom_bar(stat = "count") + xlab("Count")
# with colours - note the use of the fill in the mapping aesthetics  
ggplot(data_gb, aes(y = fct_infreq(factor(Region)), fill = Region)) + 
  geom_bar(stat = "count") + xlab("Count")
```

A final example uses bespoke colours passed to the `scale_fill_manual` functions and a brewer palette to show the number of LADs in each region:
```{r ch2fig5, eval = T, fig.cap = "An ordered barplot with bespoke shading."}
ggplot(data_gb, aes(y = fct_infreq(factor(Region)), fill = Region)) + 
  geom_bar(stat = "count") + xlab("Count") +
  scale_fill_manual("Region", values = brewer.pal(11, "Set3")) + 
  xlab("Count")
```

You can examine the range of Brewer palettes with the code below and examine them here: https://colorbrewer2.org/
```{r eval = F}
display.brewer.all()
```
You may wish to make the plot window in your RStudio session bigger to see these clearly. 
:::: {.qbox}
::: {.center}
**Task 1**
:::

The code below creates a scatter plot of the `share_leave` and `single_ethnicity_household` variables using the generic `plot` function. Your task is to recreate this using `ggplot` and add a linear trend line.
```{r eval = F}
# Task 1
plot(single_ethnicity_household ~ share_leave, data = data_gb,
     pch = 4, col = "dodgerblue", cex = 0.5)
```

::::


### Groups with `ggplot`

An important consideration is the different ways that `ggplot` can be used to display groups of data. Groups might be different treatments, classes or other discrete categories within the data, or groups that are defined by the user. There are two basic plotting options for displaying groups. 

First multiple groups can be displayed in the same plot if they are specified within the mapping aesthetics, using parameters like `size`, `colour`, `shape`, etc. The code above for bar plots did this using  the `fill` aesthetic. The code below does this with `colour` as the main grouping parameter. Notice how, having been grouped, the `geom_smooth` function plots a trend line for each group.

```{r eval = F}
p = ggplot(data = data_gb, 
       mapping = aes(x = share_leave, y = degree_educated, colour = Region)) +
  geom_point()
# have a look
p
# specify a trend line
p +  geom_smooth(method = "lm") 
```

The second option is to generate individual plots for each group using some kind of faceting. The code below generates separate scatterplots for each Region.

```{r ch2fig6, eval = T, fig.cap = "A faceted set of scatterplots."}
ggplot(data = data_gb, mapping = aes(x = share_leave, y = no_car)) +
  geom_point() +
  # add a trend line
  geom_smooth(method = "lm", colour = "#DE2D26") +
  # specify the faceting and a theme / style
  facet_wrap("Region", nrow = 3) 
```

:::: {.qbox}
::: {.center}
**Task 2**
:::
Complete the code snippet below to create a set of histograms of the `share_leave` variable, faceted for the Regions
```{r eval = F}
# Task 2
ggplot(data = data_gb, mapping = aes(x = share_leave)) +
   geom_histogram(aes(y=after_stat(density)), bins = 15, 
                 fill="tomato", col="white") + 
  # specify the faceting in the line of code below
  xxxxx 
```
::::

\

:::: {.qbox}
::: {.center}
**Task 3**
:::
The above code creates a faceted plot of the `share_leave` and  `no_car` variables, for each Region. The facets show that the relationship is broadly the same (examine the angle of the slopes) in the different regions. Your task is to examine the relationship with between the `share_leave` and `single_ethnicity_household` variables and to determine whether there are regional differences in this relationship:
::::



## Data wrangling with `dplyr`

The `dplyr` package is part of the `tidyverse`. It provides a suite of tools for manipulating and summarising data tables and spatial data and contains a number of functions for data transformations.  


A simple example is `filter` - this simply selects a subset of rows from a data frame.

The code below converts the `data_gb` to a tidy data frame and then finds the subset of data for the London Region:

```{r eval = F}
# extract the data frame
data_df = st_drop_geometry(data_gb)
# subset and assign to tmp
tmp = filter(data_df, Region == "London")
# have a look at the first 6 rows
head(tmp) 
```

```{r echo = F}
data_df = st_drop_geometry(data_gb)
```

A few things to note here:

  1. The first argument of `filter` is the data frame;
  2. The second argument is a logical expression specifying rows to be subsetted;
  3. Column names in the data frame can be used in the logical expression more or less as though they were ordinary variables;

The logical expression can be any valid logical expression in R - thus `>`,`<`,`<=`,`>=`,`==` and `!=` may all be used,  as can `&` (and), `|` (or) and `!` (not).   The `&` operator is not always necessary,  as it is possible to apply multiple conditions in the same filter operation by adding further logical arguments:

```{r eval = F}
filter(data_df, Region == "London", share_leave > 0.5)
```

The `filter` function subsets data table rows, and the `select` function subsets data table columns. This can be done in different ways as is shown in the vignette (details below) and in Part 4 of this practical. The code below simply selects by variable name

```{r eval = F}
tmp <- select(data_df, lad15nm, share_leave, total_pop)                 
head(tmp)
```

The code below uses `mutate` to create a new variable, in this case a ratio (`RL_ratio`) and assigns the result to a temporary file (`tmp`) which can be examined:

```{r eval = F}
tmp = mutate(data_df, RL_ratio = Remain/Leave) 
head(tmp)
```

Another data transform offered by `dplyr` is `arrange` which sorts the data in the order of one of the variables.   
```{r eval = F}
# ascending
head(arrange(data_df,share_leave))
# descending
head(arrange(data_df,desc(share_leave)))
```

The `filter` and `arrange` functions are just 2 of the `dplyr` *verbs* (the package developers call them verbs because they *do* something!). These can be used to manipulate data singly or in a nested a sequence of operations with intermediate outputs. The important `dyplyr` verbs are summarised in Table 1. 

```{r echo = F, message = F, warning = F}
Verb = c("filter()", "slice()", "arrange()", "desc()", "select()", "distinct()", "mutate()", "transmute()", "summarise()", "sample_n()", "sample_frac()") 
Description = c("Selects a subset of rows in a data frame according to user defined conditional statements", "Selects a subset of rows in a data frame by their position (row number)", "Reorders the row order according to the columns specified (by 1st, 2nd and then 3rd column etc)", "Orders a column in descending order", "Selects the subset of specified columns and reorders them vertically", "Finds unique values in a table", "Creates and adds new columns based on operations applied to existing columns e.g. NewCol = Col1 + Col2", "As select but only retains the new variables", "Summarises values with functions that are passed to it", "Takes a random sample of table rows", "Selects a fixed fraction of rows") 
df.tab <- data.frame(Verb, Description)

#knitr::kable(df, 
#  caption = 'The single-table verbs for manipulating data in `dplyr`.',
#  booktabs = TRUE
#)
library(kableExtra) 
knitr::kable(df.tab, 
  caption = 'The single-table verbs for manipulating data in `dplyr`.',
  booktabs = TRUE) %>% 
  kable_styling(full_width = F) %>% column_spec(2, width = "10cm") 
rm(df.tab)
```

As well as `select` (data frame columns), `filter` (rows) and `mutate` (new variables) another very useful `dplyr` function is `group_by`. This is often used in conjunction with `summarise` to create group summaries. The code below generates a summary of the population and the votes:
```{r}
# overall summary
summarise(data_df, TotPop=sum(total_pop),TotVotes=sum(Valid_Votes))
```

Note that `dplyr` manipulations can also be combined. The code below calculates population density, creating a variable called `RL_ratio` and assigns the result to `tmp1`. It then takes `tmp1` and selects this new variables with the `lad15nm` variable (the LAD name) and assigns the result to `tmp2`: 

```{r}
tmp1 = mutate(data_df, RL_ratio = Remain/Leave)
tmp2 = select(tmp1, lad15nm, RL_ratio)
head(tmp2)
```

Now the creation of all of these intermediary variables (`tmp1`, `tmp2`, etc) is a bit of pain and generates needless objects in our R session.

What we would like to be able to do is to do all of the mutating and selecting steps in one go. One option is to *nest* the commands and the code below does the same as the above but in a *nested* way: 
```{r eval = F}
tmp = select(
  mutate(data_df, RL_ratio = Remain/Leave),
  lad15nm, RL_ratio)
head(tmp)
```

Similarly the summarise operation above can be extended to work with grouped Regions:
```{r eval = F}
# regional summary
summarise(
  group_by(data_df, Region),
    TotPop=sum(total_pop),TotVotes=sum(Valid_Votes) 
  )
```

But this is getting complicated! And all we are doing here is trying to do 2 things - what if we wanted to undertake 10 operations on the data? Either we would have to create up to 10 intermediary variables or have a massively complicated nested code snippet, with nested operations as in the `summarise` and `group_by` example above. Another option is to use a *piping syntax* in which the results of one  operation are directly passed or piped to another operation. This is covered in the last section of this practical. 

After the practical you should explore the introductory vignette to `dplyr` which describes the single table verbs in `dplyr` :

```{r eval = F}
vignette("dplyr", package = "dplyr")
```

### Summary
So from the above section you should have noted that the `dplyr` package has a number of *verbs* for data table manipulations. These can be used to *select* variables or fields and to *filter* records or observations using logical statements. They can also be used to create new variables using the `mutate` function and group summaries can be easily generated using a combination of the `summarise` and `group_by` functions. The last part of the previous section highlighted that sequences of *nested* `dplyr` verbs can be applied in piped sequence rather than a clumsy code block and piped directly to `ggplot`. In the next  section we will undertake a few more manipulations of data using **piped** commands. These illustrate some key approaches for data wrangling or transformations. 

## Using piping syntax

Data in `tibble` format are basically data frames with a few modifications to help them work better in a `tidyverse` framework. A standard data frame can be converted to a tibble with the function `as_tibble` :

```{r eval = F}
data_tb <- as_tibble(data_df)
data_tb
```

```{r echo = F}
data_tb <- as_tibble(data_df)
```

Notice the difference in the print out between the `tibble` format and standard `data.frame`.


Some of the nested `dplyr` verbs and operations are difficult to follow as each `dplyr` operation required an input dataset as its first argument. When there are several layers of parenthesis (brackets), it can be hard to see immediately which parameters relate to the individual `dplyr` verbs for sorting, filtering, summarising, grouping and arranging (see for example the regional summary in the last section, reproduced below:
```{r eval = F}
# regional summary
summarise(
  group_by(data_df, Region),
    TotPop=sum(total_pop),TotVotes=sum(Valid_Votes) 
  )
```
This can be piped in the following way
```{r eval = F}
data_df |> 
  group_by(Region) |>
    summarise(TotPop=sum(total_pop),
              TotVotes=sum(Valid_Votes) )
```

Piping presents an alternative syntax that overcomes this, using the pipeline operator `|>`. Effectively this allows a variable (usually a tidy data frame) to be placed in front of a function so that `f(x)` is replaced by `x |> f()`. So, rather than the function `f()` *being applied to* `x`, `x` is *piped into* the function. The code below recreates an early operation but with piping syntax.


```{r eval = F}
# non-piped
filter(data_tb, Region == "London", share_leave > 0.5) 
# piped
data_tb |> 
  filter(Region == "London", share_leave > 0.5) 
```

:::: {.noticebox}
**Note** piping was originally introduced with in the `magrittr` package, loaded with `dplyr` and `tidyverse`. It used `%>%` as pipeline operator. So as well as the native pipe syntax of `|>`, you will also see a notation of `%>%` in places. A good summary is at https://www.tidyverse.org/blog/2023/04/base-vs-magrittr-pipe/
::::


If there are multiple arguments for `f()` we have `f(x,y,z,...)` replaced by `x \> f(y,z,...)` - thus typically the first argument is placed to the left of the function. The 'classic' R way of doing this would be as below:

```{r eval = F}
my.result <- 
  data_tb |> 
  filter(Region == "London", share_leave > 0.5) 
```

However, for some this sits uneasily with the left-to-right flow using the piping operators as the final operation (assignment) appears at the far left hand side. An alternative R assignment operator exists, however, which maintains the left-to-right flow. The operator is `->` which is illustrated with a more complex piped sequence below:

```{r}
data_tb |>
  group_by(Region) |>
  summarise(pop_dens = sum(total_pop) / sum(st_areasha * 0.01),
            share_leave=sum(Leave)/sum(Leave, Remain)) -> regional_leave_share  
```
And you can inspect the result:
```{r eval = F}
regional_leave_share
```

The remainder of this sections illustrates the use of the piping syntax in conjunction with a deeper explanation of different `dplyr` functions and verbs.  In the lecture a great explanatory video was introduced - there is a link to at the end of this practical. If you are struggling to get your head round piping then watch this now! The next subsections illustrate a few of the more commonly used `dplyr`` verbs.

### `select`

While `filter` selects subsets of rows from a data frame, `select` selects subsets of columns. A basic selection operation works as below:

```{r eval = F}
data_tb |> 
  select(Region, share_leave, single_ethnicity_household, degree_educated)
```

This selects the columns `Region`,`share_leave`, `single_ethnicity_household`, `degree_educated`. Note also that ordering of columns matters:

```{r eval = F}
data_tb |> 
  select(Region, degree_educated, share_leave, single_ethnicity_household)
```

It is possible to specify ranges of columns using `:` 

```{r eval = F}
names(data_tb)
data_tb |> 
  select(Region:share_leave)
```
Using a minus sign in front of a variable causes it to be *omitted* -

```{r eval = F}
regional_leave_share |> 
  select(-pop_dens)
```

There are also operations allowing selections based on the text in the column name. These are `starts_with`, `ends_with`, `contains` and `matches`. For example 

```{r starts, eval = F}
data_tb |> 
  select(starts_with("p"))
```

These are useful when a group of column names have some common pattern - for example `Age 0 to 5`, `Age 6 to 10` and so on could be selected using `starts_with("Age")`. The `matches` operator is more complex, as it uses regular expressions to specify matching column names. For example, suppose the data also had a column called `Agenda` - this would also be selected using `starts_with("Age")` - but using `matches("Age [0-9]")` would correctly select only the first set of columns. There is also a function called `num_range` which selects variables with a common stem and a numeric ending - e.g. `num_range("Type_",1:4)` selects `Type_1`,`Type_2`,`Type_3` and `Type_4`.

A quick note on `select`: lots of packages have a function named `select` including for example `Raster`, `dplyr`, `MASS` etc). This is the most common cause of a a select operation not working! If you have multiple packages containing a function named `select`, then you need to tell R which one to use using the double colon syntax (`::`). An example of doing this is below:

```{r eval = F}
data_tb |> 
  dplyr::select(lad15nm, Region, share_leave)
```

### `mutate` 

The `mutate` function can be used to transform variables in a data frame, or derive new variables. For example to compute the logarithm (to base 10) of the `Electorate` column in `data_df`, use

```{r eval = F}
data_tb |> 
  mutate(LogElectorate=log10(Electorate)) |> 
  select(lad15nm, Region, Electorate, LogElectorate)
```

To actually *overwrite* the `Income` column enter:

```{r eval = F}
data_tb |> 
  mutate(Electorate=log10(Electorate)) 
```

Often it is a good idea to avoid overwriting unless the results are easy to undo - otherwise it becomes very difficult to roll back if you make an error. Here, no long term damage was done since although the `mutate` operation was carried out and a tibble object was created with a redefined `Electorate` column, the new tibble wasn't stored anywhere - and most importantly it wasn't overwritten onto `data_tbl`.

A `mutate` expression can involve any of the columns in a tibble. Here, a new column containing income per head of population is created:

```{r eval = F}
data_tb |> 
  mutate(`Pop Density`= total_pop / (st_areasha *.01)) |>
  select(lad15nm, total_pop, `Pop Density`)
```

The pipelining idea can be utilised to advantage here: to create the *pop density* column and then sort on this column, reordering the columns to put the LAD name columns to the right, storing the final result into a new tibble called `new_tib` enter:

```{r tester}
data_tb |> 
  mutate(`Pop Density`= total_pop / (st_areasha *.01)) |>
  arrange(desc(`Pop Density`)) |>
  select(lad15nm, total_pop, `Pop Density`) -> new_tib
new_tib
```

A related operation is `transmute` - this is similar to `mutate` but while `mutate` keeps the existing columns in the result alongside the derived ones,`transmute` *only* returns the derived results. For example:

```{r eval = F}
data_tb |> 
  transmute(LogElectorate=log10(Electorate), LogPop=log10(total_pop))
```

To keep some of the columns from the original, but without transforming them, just state the names of these columns:

```{r}
data_tb |> 
  transmute(Region,lad15nm, LogElectorate=log10(Electorate),
            LogPop=log10(total_pop))
```

### `group_by` and `summarise`: changing the unit of observation 

The `data_tb` variable has rows corresponding to LADs, and the columns are variables (statistics) relating to each area. However, there are other geographical units in the data - as specified in the `Region` column.  There are times when it may be useful to use higher level groupings such as `Region` as a unit of observation, and aggregate variables to these units. This is where `group_by` and `summarise` are helpful. 

A key idea here is that while `mutate` carries out operations that work row-by-row on columns - so that for example `mutate(logE = log10(Electorate)` creates a column whose individual elements are the log (base 10) of the corresponding element in the column `Electorate`, `summarise` works with functions returning a single number for the whole column - such as `mean`, `median` or `sum`. Thus, for example:

```{r sum}
data_tb |> 
  summarise(TotPop=sum(total_pop),TotVotes=sum(Valid_Votes))
```

In this case the output is a tibble, but because `sum` maps a column onto a single number there is only one row, with elements corresponding to the two summarised columns. Essentially, because the sums are for the whole of the UK, the unit of observation is now the UK as a country.

However, we may wish to obtain the sums of these quantities by region. This is where `group_by` is useful. Applying this function on its own has little noticeable effect:

```{r eval = F}
data_df |> 
  group_by(Region)
```

The only visible effect is that when the tibble is printed it records that there are groups by region. Internally the variable specified for grouping is stored as an attribute of the tibble. However if this information is then passed on to `summarise` then a new effect occurs. Rather than applying the summarising expressions to a whole column, the column is split into groups on the basis of the grouping variable, and the function applied to each group in turn.

Again, pipelining is a useful tool here - a `group_by` is applied to a tibble, and the grouped tibble is piped into a `summarise` operation.
```{r}
data_df |> 
  group_by(Region) |>
  summarise(TotPop=sum(total_pop),TotVotes=sum(Valid_Votes))
```

The output is still a tibble but now the rows are the regions. As well as the summary variables `TotPop` and `TotVotes` the grouping variable (`Region`) is also included as a column. 

It is also possible to group by more than one variable. For example a new logical variable called `Dense` is introduced here based on the state population density - the population divided by the area. More rural states will have a lower density. Here `Dense` is an indicator as to whether the density exceeds 0.15. Using this and `Region` as grouping columns, the median vote leave share for each group is compared, where a group is a specific combination of each of the grouping variables. Note the use of `n()` as a summary function. This provides the number of observations in a group - here assigned to the column `N` in the output tibble.

```{r}
data_tb |> 
  mutate(Dense=total_pop/(st_areasha*0.01)> 0.15) |> 
  group_by(Region, Dense) |> 
  summarise(`Leave Share`=median(share_leave), N=n())
```

Another thing to see here is that there are 21 rows of data, but actually 22 possible `Region`/`Dense` combinations. However a row is only created if a particular combination of the grouping variables occurs in the input data. In this case, there is no un-dense area in the 'London' region, while all the other combinations do occur, hence the discrepancy of one row.

:::: {.qbox}
::: {.center}
**Task 4**
:::
Your task is to write some code that takes `data_tb`, create a new variable called `Cars` that partitions the data around `private_transport_to_work` greater than 0.5, and summarises the data by Regions and calculates the mean `leave_share`. Q: What do you think this is showing?

::::


Although perhaps the most usual use of the `group_by()` function is with `summarise()`, it is also possible to use it in conjunction with `mutate()` (and `transmute()`) and `filter()`. For `mutate` this is essentially because it is possible to use column-to-single value functions (such as `mean`,`sum`, `max` and so on) as part of an expression. Suppose we might wish to standardise the leave share votes values against a national mean figure. 
```{r eval = F}
data_tb |> 
  transmute(lad15nm,sls = 100*`share_leave`/mean(share_leave))
```

A `transmute` operation is still sensible here because the summary statistic is the denominator in a one-to-one row expression, and the overall result is still one-to-one. The numbers themselves are standardised so that very quickly one can tell whether a particular state has a life expectancy above or below the national average. For example Bromley is a little below and Bexley is a little above

However suppose we wished to standardise against regional averages rather than nation wide. If a tibble has been grouped, then operations like `sum` and so on apply to the group that each row belongs to rather than the entire data set. Thus, to standardise at the regional level, the following can be used.

```{r message=FALSE, eval = F}
data_tb |> 
  group_by(Region) |>
  transmute(lad15nm,sls = 100*`share_leave`/mean(share_leave))
```

Now the values have changed - although Barking and Dagenham still has a standardised leave voting above 100, this is more marked when considered at the regional level. 

One of the advantages of piping is that results can piped directly into `ggplot` in place of their data arguments. 

The code snippet creates a lollipop plot of the leave vote share for the different regions, without creating any intermediary variables (such as `tmp`). 


```{r ch2fig7, eval = T, fig.cap = "A lollipop plot of leave vote share."}
data_tb |>
  # group by Region
  group_by(Region) |>
  # calculate pop_dens and share_leave
  summarise(pop_dens = sum(total_pop) / sum(st_areasha * 0.01),
            share_leave=sum(Leave)/sum(Leave, Remain)) |> 
  # pass to ggplot
  ggplot(aes(x=share_leave, y=fct_reorder(Region, share_leave))) + 
  geom_point(stat='identity', fill="black", size=4)  +
  geom_segment(aes(y = Region, x = 0, yend = Region, 
                   xend = share_leave), color = "black") +
  labs(y = "Ordered Region",  x = "Leave vote share") +
  theme(axis.text=element_text(size=10))
```

The above code may appear confusing, but each line before the line with `ggplot` can be run to see what the code is doing with the data. For example try running the first few lines to unpick what is going on:
```{r}
data_tb |>
  # group by Region
  group_by(Region) |>
  # calculate pop_dens and share_leave
  summarise(pop_dens = sum(total_pop) / sum(st_areasha * 0.01),
            share_leave=sum(Leave)/sum(Leave, Remain)) 
```


### A final example

All the approaches until now have examined graphics showing univariate or bivariate (pair-wise) structure or relations. It it sometimes useful to examine multiple variables together. This can be done using *radar* plots using the `coord_polar` function with an aggregate function can be used to generate and display summaries (such as means) of the variables. The code below examines differences in the properties of LADs that had a majority Leave vote. You may also wish to examine the intermediary outputs (for example the effects of the selection, data lengthening, etc) by running blocks of code, remembering to either omit the last pipe or in this case by inserting `head()` after each `|>` (as the input is a `data.frame`).

```{r ch2fig8, eval = T, fig.cap = "A polar plot of the regional differences associated with LADs with the leave  share > 50% .", fig.width = 10}

# hard recode the region name
data_tb$Region <- recode(data_tb$Region, "Yorkshire and The Humber" = "Yorks & Humber")

data_tb %>%  
  # select the data 
  select(Region, share_leave:eu_born) |> 
  select(-total_pop, -margin_leave, -private_transport_to_work, -christian, - eu_born) |>
  # rename the variables 
  rename(leave = share_leave,
         young = younger_adults,
         engspeak = english_speaking,
         singleeh = single_ethnicity_household,
         health = not_good_health,
         degree = degree_educated,
         prof = professionals,
         nocar = no_car,
         ownhome = own_home) |>
  # rescale using z-scores for the numeric variables
  mutate_if(is.numeric, scale) |>
  # filter for Leave LADs (majority leave vote) 
  filter(as.vector(leave) > 0.5) |> 
  # now remove the leave vote 
  select(-leave) |> 
  # aggregate by region
  group_by(Region)  |>
  summarise_all(mean) |>
  # lengthen to a tidy data frame
  pivot_longer(-Region) |> 
  # sort by the variable names - this is needed for the plot
  arrange(name) |>
  ggplot() + 
  # make custom panel grid with lines
  geom_hline(
    aes(yintercept = y), 
    data.frame(y = c(-0:1)),
    color = "lightgrey"
  ) + 
  # add a dashed line for 0
  geom_hline(
    aes(yintercept = y), 
    data.frame(y = 0),
    color = "black",
    linetype = "dashed"
  ) +
  # add bars to represent the values
  geom_col(
    aes(
      x = name,
      y = value,
      group = Region, 
      fill = Region
    ),
  ) +
  # facet
  facet_wrap(~Region, ncol = 5)+
  # make polar / radar
  coord_polar() +
  # tidy the plot
  theme_light() +
  theme(legend.position = "none", 
        axis.title = element_blank(),
        axis.text = element_text(size = 8), 
        strip.text = element_text(size = 14, face = "bold", colour = "black"), 
        strip.background = element_rect(fill="white", colour = "black"), 
        panel.grid = element_line(color = "gray80", linewidth = 0.5))

```

Here we can see that there are no LADS in Scotland that had a majority Leave vote and that there are clear differences in the leave vote in different Regions. Compare for example London and Wales... there are similarities as well for example between East and the East Midlands...




## Answers to Tasks{-}

**Task 1**: the code below creates a scatter plot of the `share_leave` and `single_ethnicity_household` variables using the generic `plot` function. Your task is to recreate this using `ggplot` and add a linear trend line.
```{r eval = F}
# Task 1
plot(single_ethnicity_household ~ share_leave, data = data_gb,
     pch = 4, col = "dodgerblue", cex = 0.5)
```

```{r eval = F}
ggplot(data_gb, aes(x = share_leave, y = single_ethnicity_household)) +
  geom_point(shape = 4, col = "dodgerblue", size = 1.5) +
  geom_smooth(method = "lm") +
  theme_bw()
```
You may also want to look at at the following link for some further tips
https://felixfan.github.io/ggplot2-remove-grid-background-margin/


**Task 2**: Complete the code snippet below to create a set of histograms of the `share_leave` variable, faceted for the Regions
```{r eval = F}
# Task 2
ggplot(data = data_gb, mapping = aes(x = share_leave)) +
   geom_histogram(aes(y=after_stat(density)), bins = 15, 
                 fill="tomato", col="white") + 
  # specify the faceting
  facet_wrap("Region", nrow = 3) 
```

**Task 3**: The above code creates a faceted plot of the `share_leave` and  `no_car` variables. The facets show that the relationship is broadly the same (examine the angle of the slopes) in the different regions. Your task is to examine the relationship with between the `share_leave` and `single_ethnicity_household` variables and to determine whether there are regional differences in this relationship:

The code below explores this and the differences in the slopes show that this relationship varies across regions. Compare London (strong positive linear relationship) with the East Midlands for example (weaker flatter relationship): in this case it would seem that the association of this factors with Leave voting is much weaker in the East Midlands indicating the importance of other factors.  
```{r eval = F}
# Task 3
ggplot(data = data_gb, mapping = aes(x = share_leave, y = single_ethnicity_household)) +
  geom_point() +
  # add a trend line
  geom_smooth(method = "lm", colour = "#DE2D26") +
  # specify the faceting and a theme / style
  facet_wrap("Region", nrow = 3) +
  theme_minimal() 
```

**Task 4**: Your task is to write some code that takes `data_tb`, create a new variable called `Cars` that partitions the data around `private_transport_to_work` greater than 0.5, and summarises the data by Regions and calculates the mean `leave_share`. Q: What do you think this is showing?

```{r eval = F}
# Task 4
data_tb |> 
  mutate(Cars=private_transport_to_work> 0.5) |> 
  group_by(Region, Cars) |> 
  summarise(`Leave Share`=median(share_leave), N=n())
```

A: That areas with high / low levels of `private_transport_to_work` have very different vote leave vote shares.

## Additional Resources{-}

Density histograms explained: https://homepage.divms.uiowa.edu/~luke/classes/STAT4580/histdens.html

The BBC's use of `ggplot2`: https://medium.com/bbc-visual-and-data-journalism/how-the-bbc-visual-and-data-journalism-team-works-with-graphics-in-r-ed0b35693535

An excellent recent online book by one of our colleages: Roger Beecham (2025). Visualization for Social Data Science, Chapman and Hall/CRC Press. Free online version here: https://vis4sds.github.io/vis4sds/

There is an excellent video on piping here, although with the `dplyr` pipe of `%>%` - the message is still same: https://www.youtube.com/watch?v=ANMuuq502rE 

There is a more in-depth visualisation practical from Roger Beecham developing the  analyses of these data. You may wish to come to this later in the module after your are more familiar with R  http://homepages.see.leeds.ac.uk/~georjb/geocomp/03_datavis/lab.html

Finally: much of practical was drawn from Chapters 2, 3 and 5 of this book - https://uk.sagepub.com/en-gb/eur/geographical-data-science-and-spatial-data-analysis/book260671

EDA with R by Wickham - https://r4ds.had.co.nz/exploratory-data-analysis.html

```{r echo = F, eval = F}
knitr::purl("02-EDA.Rmd", output = "week2_ch3.R", documentation = 0)
bookdown::publish_book(name = "GEOG3195", account = "lexcomber")
```

There are some useful Data Resources also:

- ONS spatial data: https://geoportal.statistics.gov.uk/search?collection=Map
- Spatial data from ONS (shapefile format) - https://www.data.gov.uk/search?q=&filters%5Bpublisher%5D=&filters%5Btopic%5D=Mapping&filters%5Bformat%5D=SHP&sort=best
- House of Commons data: https://commonslibrary.parliament.uk/data-tools-and-resources/
- uk-hex-cartograms-noncontiguous, House of Commons github, https://github.com/houseofcommonslibrary/uk-hex-cartograms-noncontiguous
- the `parlitools` package has lots of voting data https://cran.r-project.org/web/packages/parlitools/vignettes/introduction.html



## Optional: enhanced mapping with the `ggplot2` package {-}

As you saw on Week 1, the `ggplot2` package can also be used to visualise spatial data. This section provides some further examples of how to do this with point and area data. 

You will need to load  the `ggspatial` package into your R session: 

```{r}
# ggspatial for mapping with ggplot
if (!is.element("ggspatial", installed.packages()))
    install.packages("ggspatial", dep = T)
library(ggspatial)
```

The code below selects the LADs in the East Midlands for us to work with to create an `sf` object from a subset of `data_gb`:

```{r warning=F, message=F}
east_mids = data_gb[data_gb$Region == "East Midlands", ]
```
The code below plots the `east_mids` dataset introduced above using the `ggplot2` package: 

```{r ch5fig15, cache = F, eval=TRUE, echo=T, message=FALSE, fig.cap = "The LADs in the East Midlands."}
ggplot(east_mids) + geom_sf()
```

The `geom_sf` function, adds a geometry stored in an `sf` object to the plot. It looks for an attribute called `geom` or `geometry` in the `sf` object. On occasion you may need to specify this as in the code snippet below.
```{r eval = F}
ggplot(east_mids) + geom_sf(aes(geometry = geometry))
```

As with  other `ggplot` operations, colour can be specified directly or for a variable, by specifying plotting aesthetics (`aes`) using `fill` or `colour` depending on the what is being shaded in the plot:
```{r eval = F}
# directly
ggplot(east_mids) +
  geom_sf(aes(geometry = geometry), colour = "darkgrey", fill = "azure")
# using a continuous variable and specifying a thin line size
ggplot(east_mids) +
  geom_sf(aes(geometry = geometry, fill = degree_educated), linewidth = 0.1) +
  scale_fill_viridis_c(name = "Degree educated")
```

It is also possible to **enhance** the map by adding scale bars and north arrows using some of the functions in the `ggspatial` package. We can also control the appearance of the map using both predefined themes (such `theme_bw`) and specific `theme` elements. Examples of some of these are shown in the code below to create the figure below:

```{r ch2fig17, cache = F, eval=TRUE, echo=T, message=FALSE, fig.cap = "Poor Health in LADs in the East Midlands."}
ggplot(east_mids) +
  geom_sf(aes(geometry = geometry, fill = not_good_health), size = 0.1) +
  scale_fill_viridis_c(name = "Poor Health") + 
  annotation_scale(location = "tl") +
  annotation_north_arrow(location = "tl", which_north = "true", 
        pad_x = unit(0.9, "in"), pad_y = unit(0.25, "in"),
        style = north_arrow_fancy_orienteering) +
  theme_bw() +
  theme(panel.grid.major = element_line(color = gray(.5), linetype="dashed", linewidth=0.5),
        panel.background = element_rect(fill="white"))
```

Multiple data layers can be included in the map. The code below creates and then adds the locations and names of some randomly selected LADs to the map (**Note** the `set.seed()` function makes sure that you all select the same random sample!).
```{r eval = F}
set.seed(13)
labels.data = east_mids[sample(1:nrow(east_mids), 10), ]
labels.data = cbind(labels.data, st_coordinates(st_centroid(labels.data)))
# create the ggplot
ggplot(east_mids) +
  geom_sf(aes(geometry = geometry), fill = "white", linewidth = 0.1) +
  geom_point(data = labels.data, aes(x=X, y=Y), colour = "red") +
  geom_text(data = labels.data, aes(x=X, y=Y, label=lad15nm),
    fontface = "bold", check_overlap = T) +
  theme_bw() +
  theme(axis.title = element_blank())
```

A final trick is to include a backdrop to the map. This is not interactive, but instead loads tiles of backdrop maps to provide context. This can be done using the `annotation_map_tile` function from `ggspatial`. Note the order it needs to be placed in the sequence `ggplot` layers of commands:
```{r ch2fig18, warning=F, message=F, fig.cap = "An example of using OSM as background context."}
ggplot(east_mids) +
  annotation_map_tile(zoomin=0, type='osm') + 
  geom_sf(aes(geometry = geometry, fill = degree_educated), linewidth = 0.1, alpha = 0.5) +
  scale_fill_viridis_c(name = "Degree educated")
```

Often in geographic analyses were are interested **where** behaviour changes or flips.  So for example a key thing we may be interested in, that cannot be directly determined using the *aspatial* bivariate and multivariate approaches (correlations, correspondence tables etc), are the locations *where* for example leave votes  were high or low.

The code below partitions the data around median leave share. This shows that there are clear spatial trends. 
```{r ch2fig19, fig.height = 10, cache = F, eval=TRUE, echo=T, results='hide',message=FALSE, fig.cap = "A map of LADs where the leave vote share was higher and lower than the median."} 
data_gb %>% 
  mutate(leave = ifelse(share_leave > median(share_leave), "Higher","Lower")) %>%
  ggplot() + 
    geom_sf(aes(geometry = geometry, fill = leave), col = "grey", size = 0.0) +
    scale_fill_discrete(name = "Leave > median?") +
    theme_bw()

```    


## Appendix: Named Colours{-}

This section is just for your information. 

R has a number of named colours. You can list them by entering
```{r eval = F}
colours()
```

For those interested the code used is listed below. You’ll need the `kableExtra` package installed to get this to work.

```{r, eval = F}
library(kableExtra)

### Matrix of rgb values for each named colour
rgbmat <- t(col2rgb(colors()))

### Turned into hex formats
hexes <- apply(rgbmat, 1, function(x) sprintf("#%02X%02X%02X",x[1],x[2],x[3]))

### These will be backgrounds - foreground either black or white depending on lightness of colour
textcol <- ifelse(rowMeans(rgbmat) > 156, 'black', 'white')

### Now create the cells with fg/bg colour spec
cols <- colours()
cols <- cell_spec(cols,background=hexes,color=textcol)

### Make this a grid
colour_grid <- matrix(c(cols,rep("",7)),83,8)
colnames(colour_grid)=rep(" ",8)

### Use kableExtra stuff to create html table
ktab <- kable(colour_grid,align='l',format='html',escape=FALSE)
kable_styling(ktab)
```

```{r echo = F, eval = F}
knitr::purl("01-Piping_EDA.Rmd", output = "week2.R", documentation = 0)
bookdown::publish_book(name = "GEOG3195", account = "lexcomber")
```
