# Creating documents in RMarkdown

## Overview {-}

In this practical you will create an RMarkdown (`.Rmd`) document in either `html` or `PDF` format. It can include anything, but the following things are included in RMarkdown document on the VLE (`Rep_example.Rmd`):

- text in paragraphs
- plots (e.g. scatterplots)
- maps
- tables
- some hidden code 
- some visible code

**NOTE** 

1) you may have to install the `rmarkdown` package on University computers - you should not have to but  you never know! You should test this by running the code:
```{r eval = F}
if (!is.element("rmarkdown", installed.packages()))
    install.packages("rmarkdown", dep = T)
```

2) You may need to use a simply named folder. This means removing any non-text characters in the name. So 

- **not** C:\\Geocomp & Loc Anal\\week9
- **instead** C:\\Geocomp\\week9

## Resources

This tutorial https://rmarkdown.rstudio.com/articles_intro.html may help and some other potentially useful resources include:

- https://www.neonscience.org/rmd-code-intro
- https://rpubs.com/mnewco8290/94529
- http://www.jacolienvanrij.com/Tutorials/tutorialMarkdown.html

Also there are Cheatsheets 

- https://rmarkdown.rstudio.com/lesson-15.html
- they can also be accessed through the Rstudio menu **Help > Cheat Sheets > R Markdown Cheat Sheet**

Documents created in this way allow you to connect the code, the analysis and the data in a single environment, supporting transparency and reproducibility. 

## Tasks

**Task 1**

This just gets you to create,  save and compile a simple R Markdown document.

1. Open a new RMarkdown document:

**File > New File > R Markdown**

use HTML format.

Give it a name, save it with a name in your working folder, and try to run it by clicking on the "knit" icon. If you get an error here, you probably need to install the `rmarkdown` package for some reason.

2. Then change the 3rd line from 

- from `output: html_document`
- to `output: pdf_document`

and compile the document again to create a PDF file.
**NOTE** that if you are on a University PC you may get a 'cannot compile PDF' error message. In which case you may need to install a PDF compiler.
```{r eval = F}
# only run this if you get the above errror!
tinytex::install_tinytex()
```

**Task 2**

Go through the Rep_Example.Rmd files on the VLE. I have chopped down a GWR Practical as a short example in the RMD file. You should 

- make sure that you can identify sections of words and code that produce text, plots, maps, hidden / visible code and tables as indicated above. 
- pay attention to the way that the code snippets presented, especially how the code blocks are bracketed by the backticks ( ` ).
- try to create both PDF and HTML outputs by changing the outputs as below in the RMD header:
  - `output: pdf_document`
  - `output: html_document`


```{r echo = F, eval = F}
bookdown::publish_book(name = "GEOG3195", account = "lexcomber")
```







